### 2018-06-30 paper/3-abstractions.tex
Work on S2-S3, add a figure with key abstractions

This section presents purely functional abstractions that allow us to express
all the intricacies of build systems discussed previously in \S\ref{sec-background},
and design complex build systems from simple
primitives. Specifically, we present the \emph{task} and \emph{build}
abstractions in~\S\ref{sec-task} and~\S\ref{sec-general-build}, respectively.
Sections~\S\ref{sec-build} and~\S\ref{sec-implementations} scrutinise the
abstractions further and provide concrete implementations for several build
systems.

\subsection{Common vocabulary for build systems}\label{sec-vocabulary}
\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- An abstract store
data Store i k v -- i = info, k = key, v = value
getInfo    :: Store i k v -> i
putInfo    :: i -> Store i k v -> Store i k v
getValue   :: k -> Store i k v -> v
putValue   :: Eq k => k -> v -> Store i k v -> Store i k v
getHash    :: Hashable v => k -> Store i k v -> Hash v
initialise :: i -> (k -> v) -> Store i k v
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Hashing
hash :: Hashable a => a -> Hash a
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Applicative functors
pure  :: Applicative f => a -> f a
(<$>) :: Applicative f =>   (a -> b) -> f a -> f b -- Left-associative
(<*>) :: Applicative f => f (a -> b) -> f a -> f b -- Left-associative
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard State monad from Control.Monad.State
data State s a
instance Applicative (State s)
instance Monad       (State s)
get       :: State s s
gets      :: (s -> a) -> State s a
modify    :: (s -> s) -> State s ()
execState :: State s a -> s -> s
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard types from Data.Functor.Identity and Data.Functor.Const
newtype Identity a = Identity { runIdentity :: a }
newtype Const m a  = Const    { getConst    :: m }
\end{minted}
\vspace{0.5mm}
\begin{minted}[fontsize=\small]{haskell}
instance Functor (Const m) where
    fmap _ (Const m) = Const m
\end{minted}
\vspace{0.5mm}
\begin{minted}[fontsize=\small]{haskell}
instance Monoid m => Applicative (Const m) where
    pure _              = Const mempty
    Const x <*> Const y = Const (x <> y)
\end{minted}
\vspace{-3mm}
\caption{Signatures of main data types and library functions.}\label{fig-types}
\vspace{-4mm}
\end{figure}
We begin by establishing a common vocabulary for build systems:

\vspace{-2mm}
\paragraph{Keys, values, and the store} The goal of any build system is to
bring up to date a \emph{store} that implements a mapping from \emph{keys} to
\emph{values}. In software build systems the store is the file system, the
keys are filenames, and the values are file contents. In \Excel, the store is
the worksheets, the keys are cell names (such as \cmd{A2}) and the values are
numbers, strings etc, displayed as the cell contents. Many build systems use
\emph{hashes} of values as compact summaries with a fast equality check.

\vspace{-2mm}
\paragraph{Input, output, and intermediate values}
Some values must be provided by the user as \emph{input}. For example,
\cmd{main.c} can be edited by the user who relies on the build system to
compile it into \cmd{main.o} and subsequently \cmd{main.exe}. End build products,
such as \cmd{main.exe}, are \emph{output} values. All other values (in this case
\cmd{main.o}) are \emph{intermediate}; they are not interesting for the user
but are produced in the process of turning inputs into outputs.

\vspace{-2mm}
\paragraph{Persistent build information} As well as the key/value mapping, the store
also contains information maintained by the build system itself, which persists
from one invocation of the build system to the next -- its ``memory''.

\vspace{-2mm}
\paragraph{Task description} Any build system requires the user to specify how to
compute the new value for one key, using the (up to date) values of its
dependencies. We call this specification the \emph{task description}.  For
example, in \Excel, the formulae of the spreadsheet constitute the task
description; in \Make the rules in the makefile are the task description.

\vspace{-2mm}
\paragraph{Build system} A \emph{build system} takes a task description, a
\emph{target} key, and a store, and returns a new store in which the target key
and all its dependencies have an up to date value.

\vspace{-2mm}
\paragraph{Modelling in Haskell} We will model all our build systems concretely,
as Haskell programs. To that end, Fig.~\ref{fig-types} gives the type
declarations and function signatures of the library functions. For example,
\hs{Store}~\hs{i}~\hs{k}~\hs{v} is the type of stores, with several associate
functions (\hs{getValue}, etc.). We use \hs{k} as a type
variable ranging over keys, \hs{v} for values, and \hs{i} for the persistent
build information.

@@

% Old type:
% type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{text}
A1: 10     B1: A1 + A2
A2: 20     B2: B1 * 2
\end{minted}
\vspace{1mm}

\noindent
Here cell \cmd{A1} contains the value \cmd{10}, cell \cmd{B1} contains
the formula \cmd{A1+A2}, etc. We can represent the formulae of this spreadsheet
with the following task description:

@@

The code to ``compute the value of a formula'' in \hs{sprsh1} looks a bit
mysterious because it takes place in an \hs{Applicative}
computation~\cite{mcbride2008applicative} -- the relevant type signatures are
given in Fig.~\ref{fig-types}. We will explain why in~\S\ref{sec-general-build}.

For now, we content ourselves with observing that a task description, of type
\hs{Tasks}~\hs{c}~\hs{k}~\hs{v}, is completely isolated from the world of
compilers, calculation chains, file systems, caches, and all other
complexities of real build systems.  It just computes a single output, in
a side-effect-free way, using a callback (\hs{fetch}) to find the values
of its dependencies.

@@

\noindent
The \hs{busy} build system defines the callback \hs{fetch} so that, when given a
key, it brings the key up to date in the store, and returns its value.
The function \hs{fetch} runs in the standard Haskell \hs{State} monad -- see
Fig.~\ref{fig-types} -- initialised with the incoming \hs{store} by \hs{execState}.
To bring a key up to date, \hs{fetch} asks the task description \hs{tasks} how
to compute the value of \hs{k}. If \hs{tasks} returns \hs{Nothing} the key is an
input, so \hs{fetch} simply reads the result from the store. Otherwise \hs{fetch}
runs the obtained \hs{task} to produce a resulting value~\hs{v}, records the new
key/value mapping in the store, and returns \hs{v}. Notice that \hs{fetch}
passes itself to \hs{task} as an argument, so that the latter can use \hs{fetch}
to recursively find the values of \hs{k}'s dependencies.

@@

\subsection{The need for polymorphism in \hs{Task}}\label{sec-why-polymorphism}

@@

But nor can the task description work for \emph{any} \hs{f}; most task
descriptions (e.g. \hs{sprsh1} in \S\ref{sec-task}) require that \hs{f}
satisfies certain properties, such as \hs{Applicative} or \hs{Monad}. That is
why \hs{Task} has the ``\hs{c}~\hs{f}~\hs{=>}'' constraint in its type,
expressing that \hs{f} can only be instantiated by types that satisfy the
constraint \hs{c}.

So the type \hs{Task} emerges naturally, almost inevitably.  But now that
it \emph{has} emerged, we find that constraints \hs{c} classify task descriptions
in a very interesting way:
\begin{itemize}
\item \hs{Task}~\hs{Applicative}. In \hs{sprsh1} we needed only \hs{Applicative}
  operations, expressing the fact that the dependencies between cells can be
  determined \emph{statically}; that is, by looking at the formulae, without
  ``computing'' them (see \S\ref{sec-deps}).
\item \hs{Task}~\hs{Monad}. As we shall see in \S\ref{sec-task-monad}, a monadic task
  description allows \emph{dynamic} dependencies, in which a formula may depend
  on the value of cell \cmd{C}, but \emph{which} cell \cmd{C} depends on the
  value of another cell \cmd{D}.
%TODO: Add a reference for Docker?
\item \hs{Task}~\hs{Functor} is somewhat degenerate: the task description cannot
  even use the application operator \hs{<*>}, which limits dependencies to a
  single linear chain, as e.g. in the Docker build system. It is interesting to
  note that, when run on such a task description, the \hs{busy} build system
  will build each key at most once, thus partially fulfilling the minimality
  requirement~\ref{def-minimal}. Alas, it still has no mechanism to decide which
  input keys changed since the previous build.
\item \hs{Task}~\hs{Alternative}, \hs{Task}~\hs{MonadPlus} and their
  variants can be used for describing tasks with a certain type of
  non-determinism, as discussed in~\S\ref{sec-non-determinism}.
\item \hs{Task}~\hs{(MonadState}~\hs{i)} will be used
  in~\S\ref{sec-implementations} to describe tasks that have read and write
  access to the persistently stored build information~\hs{i}.
\end{itemize}

@@

\subsection{Monadic tasks}\label{sec-task-monad}

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{text}
A1: 10      B1: IF(C1=1,B2,A2)      C1: 1
A2: 20      B2: IF(C1=1,A1,B1)
\end{minted}

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
sprsh2 :: Tasks Monad String Integer
sprsh2 "B1" = Just $ Task $ \fetch -> do
    c1 <- fetch "C1"
    if c1 == 1 then fetch "B2" else fetch "A2"
sprsh2 "B2" = Just $ Task $ \fetch -> do
    c1 <- fetch "C1"
    if c1 == 1 then fetch "A1" else fetch "B1"
sprsh2 _ = Nothing
\end{minted}
\vspace{1mm}

@@

\noindent
Here we do not need any effects in the \hs{fetch} callback to \hs{task}, so
we can use the standard Haskell \hs{Identity} monad (Fig.~\ref{fig-types}).
The use of \hs{Identity} just fixes the `impedance mismatch' between the
function \hs{getValue}, which returns a pure value~\hs{v}, and the \hs{fetch}
argument of the \hs{task}, which must return an \hs{f}~\hs{v} for some \hs{f}.
To fix the mismatch, we wrap the result of \hs{getValue} in the \hs{Identity}
monad: the function
\hs{\@@k}~\hs{->}~\hs{Identity}~\hs{(@@getValue}~\hs{k}~\hs{store)} has the type
\hs{k}~\hs{->}~\hs{Identity}~\hs{v}, and can now be passed to a \hs{task}. The
result comes as \hs{Identity}~\hs{v}, which we unwrap by \hs{runIdentity}.

@@

\noindent
Here \hs{Const} is a standard Haskell type defined in Fig.~\ref{fig-types}. We
instantiate \hs{f} to \hs{Const}~\hs{[@@k]}. So a value of type \hs{f}~\hs{v},
or in this case \hs{Const}~\hs{[@@k]}~\hs{v}, contains no value \hs{v}, but does
contain a list of keys of type \hs{[@@k]} which we use to record dependencies.
The \hs{fetch} callback that we pass to \hs{task} records a single dependency;
and the standard definition of \hs{Applicative} for \hs{Const} (which we give
in Fig.~\ref{fig-types}) combines the dependencies from different parts of the
task. Running the task with \hs{f}~=~\hs{Const}~\hs{[@@k]} will thus
accumulate a list of the task's dependencies -- and that is just what
\hs{dependencies} does:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies $ fromJust $ sprsh1 "A1"
[@@]
\end{minted}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies $ fromJust $ sprsh1 "B1"
["A1", "A2"]
\end{minted}
\vspace{1mm}

\noindent
Notice that these calls to \hs{dependencies} do no actual computation (in this
case, spreadsheet arithmetic). They cannot: we are not supplying a store or any
input numbers. So, through the wonders of polymorphism, we are able to extract
the dependencies of the spreadsheet formula, and to do so efficiently, simply by
running its code in a different \hs{Applicative}! This is not new, for example
see~\cite{free-applicatives}, but it is cool.

@@

\noindent
This implementation uses the standard Haskell \hs{WriterT} \emph{monad
transformer}~\cite{liang1995monad}, which allows us
to record additional information -- a list of keys of type \hs{[@@k]} -- when
computing a task in an arbitrary monad~\hs{m}. We substitute the given \hs{fetch}
with a \hs{trackingFetch} that, in addition to fetching a value, tracks the
corresponding key. The \hs{task} returns the value of type
\hs{WriterT}~\hs{[@@k]}~\hs{m}~\hs{v}, which we unwrap by \hs{runWriterT}. Here
is an example of \hs{track}ing monadic tasks when \hs{m}~\hs{=}~\hs{IO}.

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ fetchIO k = do putStr (k ++ ": "); read <$> getLine
@\ghci@ track (fromJust $ sprsh2 "B1") fetchIO
C1: 1
B2: 10
(10,["C1","B2"])
\end{minted}
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ track (fromJust $ sprsh2 "B1") fetchIO
C1: 2
A2: 20
(20,["C1","A2"])
\end{minted}
\vspace{1mm}

@@

% Furthermore, the above example highlights the static cycle between cells
% \cmd{B1} and \cmd{B2}: in \hs{store1}, we have \cmd{B1 = B2 = A1 = 10},
% whereas in \hs{store2}, we have \cmd{B2 = B1 = A2 = 20}.

% \subsection{Free tasks}
% \todo{AM}{Introduce free constructions \hs{TaskA} and \hs{TaskM} giving
% intuitions on what tasks really are. This will take at least one page.}
%
% data TaskA k v a = InputA | TaskA [k] ([v] -> a)
%
% data TaskM k v a = InputM | DoneM a | TaskM [k] ([v] -> Task k v a)

% \subsection{Recognising an applicative task wearing a monadic hat}
% Although dynamic dependencies are very useful, they typically represent only a
% small fraction of all dependencies. One can exploit this in a build system by
% recognising applicative tasks even though they have the type
% \hs{Task}~\hs{Monad} and treating them specially.

%------------------------- Old correctness definition -------------------------%

% \definition[Correctness]{Suppose \hs{build} is a build system, \hs{task} is a
% build task description, \hs{key} is an output key, \hs{store} is an initial
% store, and \hs{result} is the store produced by running the build system with
% parameters \hs{task}, \hs{key} and \hs{store}. Or, using the precise language of
% our abstractions:
% \vspace{1mm}
% \begin{minted}[xleftmargin=10pt]{haskell}
% build         :: Build c i k v
% task          :: Task c k v
% key           :: k
% store, @@result :: Store i k v
% result = @@build @@task @@key @@store
% \end{minted}
% \vspace{1mm}
% \noindent
% Then the \hs{result} is~\emph{correct} if there exists an \hs{ideal} store, such
% that:
% \begin{itemize}
%     \item The \hs{ideal} store is \emph{consistent} with the \hs{task}, i.e.
%     for all possible non-input keys \hs{k}, the result of recomputing the
%     \hs{task} matches the value in the \hs{ideal} store:
%     \vspace{-1mm}
%     \[
%     \hs{Just}~\hs{(}\hs{getValue}~\hs{ideal}~\hs{k)}~\hs{==}~\hs{compute}~\hs{task}~\hs{ideal}~\hs{k}.
%     \]
%     \item The \hs{result} and \hs{ideal} agree on the output \hs{key}, i.e.:
%     \vspace{-1mm}
%     \[
%     \hs{getValue}~\hs{result}~\hs{key}~\hs{==}~\hs{getValue}~\hs{ideal}~\hs{key}.
%     \]
%     In other words, we require that the output \hs{key} has the correct value,
%     but do not impose any restrictions on intermediate keys, therefore
%     permitting shallow cloud builds.
%     \item The \hs{store}, \hs{result} and \hs{ideal} agree on all input keys
%     \hs{k} that belong to the transitive closure of \hs{key}'s dependencies:
%     \[
%     \,\,\,\,\,\,\,\hs{getValue}~\hs{store}~\hs{k}~\hs{==}~\hs{getValue}~\hs{result}~\hs{k}~\hs{&&}~\hs{getValue}~\hs{store}~\hs{k}~\hs{==}~\hs{getValue}~\hs{ideal}~\hs{k}
%     \]
%     This requirement asserts that (i) no inputs were corrupted during the
%     build, and (ii)~the \hs{ideal} store has constraints not only on the output,
%     but also on the inputs.
% \end{itemize}
% A build system is \emph{correct} if it produces a correct \hs{result} for any
% given \hs{task}, \hs{key} and \hs{store}.
% }
% \label{def-correct}

>>>>

\begin{figure}
\vspace{-1mm}
\begin{minted}[fontsize=\small]{haskell}
-- An abstract store containing a key/value map and persistent build information
data Store i k v -- i = info, k = key, v = value
initialise :: i -> (k -> v) -> Store i k v
getInfo    :: Store i k v -> i
putInfo    :: i -> Store i k v -> Store i k v
getValue   :: k -> Store i k v -> v
putValue   :: Eq k => k -> v -> Store i k v -> Store i k v
getHash    :: Hashable v => k -> Store i k v -> Hash v
hash       :: Hashable v => v -> Hash v -- provides a fast equality check
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build tasks (see §3.2)
newtype Task  c k v = Task { run :: @\std{forall}@ f. c f => (k -> f v) -> f v }
type    Tasks c k v = k -> Maybe (Task c k v)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build system (see §3.3)
type Build c i k v = Tasks c k v -> k -> Store i k v -> Store i k v
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build system components: a scheduler and a rebuilder (see §5)
type Scheduler c i ir k v = Rebuilder c ir k v -> Build c i k v
type Rebuilder c   ir k v = k -> v -> Task c k v -> Task (MonadState ir) k v
\end{minted}
\vspace{-3mm}
\caption{Type signatures of key build systems abstractions.}\label{fig-types}
\vspace{-3mm}
\end{figure}

This section presents purely functional abstractions that allow us to express
all the intricacies of build systems discussed in~\S\ref{sec-background},
and design complex build systems from simple primitives. Specifically, we
present the \emph{task} and \emph{build} abstractions in~\S\ref{sec-task}
and~\S\ref{sec-general-build}, respectively. Sections~\S\ref{sec-build}
and~\S\ref{sec-implementations} scrutinise the abstractions further and provide
concrete implementations for several build systems.

\vspace{-1.5mm}
\subsection{Common vocabulary for build systems}\label{sec-vocabulary}
\vspace{-0.5mm}
% We begin by establishing a common vocabulary for build systems:

\emph{Keys, values, and the store.} The goal of any build system is to
bring up to date a \emph{store} that implements a mapping from \emph{keys} to
\emph{values}. In software build systems the store is the file system, the
keys are filenames, and the values are file contents. In \Excel, the store is
the worksheets, the keys are cell names (such as \cmd{A1}) and the values are
numbers, strings, etc., displayed as the cell contents. Many build systems use
\emph{hashes} of values as compact summaries with a fast equality check.

\emph{Input, output, and intermediate values.} Some values must be provided by
the user as \emph{input}. For example, \cmd{main.c} can be edited by the user
who relies on the build system to compile it into \cmd{main.o} and subsequently
\cmd{main.exe}. End build products, such as \cmd{main.exe}, are \emph{output}
values. All other values (in this case \cmd{main.o}) are \emph{intermediate};
they are not interesting for the user but are produced in the process of turning
inputs into outputs.

\emph{Persistent build information.} As well as the key/value mapping, the
store also contains information maintained by the build system itself, which
persists from one invocation of the build system to the next -- its ``memory''.

\emph{Task description.} Any build system requires the user to specify how
to compute the new value for one key, using the (up to date) values of its
dependencies. We call this specification the \emph{task description}. For
example, in \Excel, the formulae of the spreadsheet constitute the task
description; in \Make the rules in the makefile are the task description.

\emph{Build system.} A \emph{build system} takes a task description, a
\emph{target} key, and a store, and returns a new store in which the target key
and all its dependencies have an up to date value.

We will model build systems concretely, as Haskell programs. To that end,
Fig.~\ref{fig-types} provides the type signatures for all key abstractions
introduced in the paper. For example, \hs{Store}~\hs{i}~\hs{k}~\hs{v} is the
type of stores, with several associate functions (\hs{getValue}, etc.). We use
\hs{k} as a type variable ranging over keys, \hs{v} for values, and \hs{i} for
the persistent build information. Fig.~\ref{fig-stdlib} lists standard library
definitions.

@@

@@

\begin{minted}[xleftmargin=10pt]{text}
A1: 10     B1: A1 + A2
A2: 20     B2: B1 * 2
\end{minted}

\noindent
Here cell \cmd{A1} contains the value \cmd{10}, cell \cmd{B1} contains the
formula \cmd{A1+A2}, etc.

% \newpage

\noindent
We can represent the formulae of this spreadsheet with the following task
description:

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Applicative functors
pure  :: Applicative f => a -> f a
(<$>) :: Applicative f =>   (a -> b) -> f a -> f b -- Left-associative
(<*>) :: Applicative f => f (a -> b) -> f a -> f b -- Left-associative
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard State monad from Control.Monad.State
data State s a
instance Monad (State s)
get       :: State s s
gets      :: (s -> a) -> State s a
put       :: s -> State s ()
modify    :: (s -> s) -> State s ()
runState  :: State s a -> s -> (a, s)
execState :: State s a -> s -> s
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard types from Data.Functor.Identity and Data.Functor.Const
newtype Identity a = Identity { runIdentity :: a }
newtype Const m a  = Const    { getConst    :: m }
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
instance Functor (Const m) where
    fmap _ (Const m) = Const m
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
instance Monoid m => Applicative (Const m) where
    pure _              = Const mempty
    Const x <*> Const y = Const (x <> y)
\end{minted}
\vspace{-3mm}
\caption{Standard library definitions.}\label{fig-stdlib}
\vspace{-4mm}
\end{figure}

The code to ``compute the value of a formula'' in \hs{sprsh1} looks a bit
mysterious because it takes place in an \hs{Applicative}
computation~\cite{mcbride2008applicative} -- the relevant type signatures are
given in Fig.~\ref{fig-stdlib}. We will explain why in~\S\ref{sec-general-build}.

For now, we content ourselves with observing that a task description, of type
\hs{Tasks}~\hs{c}~\hs{k}~\hs{v}, is completely isolated from the world of
compilers, calc chains, file systems, caches, and all other complexities of real
build systems. It just computes a single output, in a side-effect-free way,
using a callback (\hs{fetch}) to find the values of its dependencies.

@@

\noindent
The \hs{busy} build system defines the callback \hs{fetch} so that, when given a
key, it brings the key up to date in the store, and returns its value.
The function \hs{fetch} runs in the standard Haskell \hs{State} monad -- see
Fig.~\ref{fig-stdlib} -- initialised with the incoming \hs{store} by \hs{execState}.
To bring a key up to date, \hs{fetch} asks the task description \hs{tasks} how
to compute the value of \hs{k}. If \hs{tasks} returns \hs{Nothing} the key is an
input, so \hs{fetch} simply reads the result from the store. Otherwise \hs{fetch}
runs the obtained \hs{task} to produce a resulting value~\hs{v}, records the new
key/value mapping in the store, and returns \hs{v}. Notice that \hs{fetch}
passes itself to \hs{task} as an argument, so that the latter can use \hs{fetch}
to recursively find the values of \hs{k}'s dependencies.

@@

\subsection{The need for polymorphism in \hs{Task}}\label{sec-why-polymorphism}
\vspace{-0.5mm}

@@

But nor can the task description work for \emph{any} \hs{f}; most task
descriptions (e.g. \hs{sprsh1} in \S\ref{sec-task}) require that \hs{f}
satisfies certain properties, such as \hs{Applicative} or \hs{Monad}. That is
why \hs{Task} has the ``\hs{c}~\hs{f}~\hs{=>}'' constraint in its type,
expressing that \hs{f} can only be instantiated by types that satisfy the
constraint \hs{c}. So the type \hs{Task} emerges naturally, almost inevitably.
But now that it \emph{has} emerged, we find that constraints \hs{c} classify
task descriptions in a very interesting way:
\begin{itemize}
\item \hs{Task}~\hs{Applicative}. In \hs{sprsh1} we needed only \hs{Applicative}
  operations, expressing the fact that the dependencies between cells can be
  determined \emph{statically}; that is, by looking at the formulae, without
  ``computing'' them (see \S\ref{sec-deps}).
\item \hs{Task}~\hs{Monad}. As we shall see in \S\ref{sec-task-monad}, a monadic
  task allows \emph{dynamic} dependencies, in which a formula may depend on cell
  \cmd{C}, but \emph{which} cell \cmd{C} depends on the value of another cell
  \cmd{D}.
%TODO: Add a reference for Docker?
\item \hs{Task}~\hs{Functor} is somewhat degenerate: the task description cannot
  even use the application operator \hs{<*>}, which limits dependencies to a
  single linear chain, as e.g. in the Docker build system. It is interesting to
  note that, when run on such a task description, the \hs{busy} build system
  will build each key at most once, thus partially fulfilling the minimality
  requirement~\ref{def-minimal}. Alas, it still has no mechanism to decide which
  input keys changed since the previous build.
\item \hs{Task}~\hs{Alternative}, \hs{Task}~\hs{MonadPlus} and their
  variants can be used for describing tasks with a certain type of
  non-determinism, as discussed in~\S\ref{sec-non-determinism}.
\item \hs{Task}~\hs{(MonadState}~\hs{i)} will be used
  in~\S\ref{sec-implementations} to describe tasks that have read and write
  access to the persistently stored build information~\hs{i}.
\end{itemize}

@@

\vspace{-0.5mm}
\subsection{Monadic tasks}\label{sec-task-monad}

@@

\vspace{0.5mm}
\begin{minted}[xleftmargin=10pt]{text}
A1: 10      B1: IF(C1=1,B2,A2)      C1: 1
A2: 20      B2: IF(C1=1,A1,B1)
\end{minted}
\vspace{0.5mm}

@@

\vspace{0.5mm}
\begin{minted}[xleftmargin=10pt]{haskell}
sprsh2 :: Tasks Monad String Integer
sprsh2 "B1" = Just $ Task $ \fetch -> do
    c1 <- fetch "C1"
    if c1 == 1 then fetch "B2" else fetch "A2"
sprsh2 "B2" = Just $ Task $ \fetch -> do
    c1 <- fetch "C1"
    if c1 == 1 then fetch "A1" else fetch "B1"
sprsh2 _ = Nothing
\end{minted}
\vspace{0.5mm}

@@

\noindent
Here we do not need any effects in the \hs{fetch} callback to \hs{task}, so
we can use the standard Haskell \hs{Identity} monad (Fig.~\ref{fig-stdlib}).
The use of \hs{Identity} just fixes the `impedance mismatch' between the
function \hs{getValue}, which returns a pure value~\hs{v}, and the \hs{fetch}
argument of the \hs{task}, which must return an \hs{f}~\hs{v} for some \hs{f}.
To fix the mismatch, we wrap the result of \hs{getValue} in the \hs{Identity}
monad: the function
\hs{\@@k}~\hs{->}~\hs{Identity}~\hs{(@@getValue}~\hs{k}~\hs{store)} has the type
\hs{k}~\hs{->}~\hs{Identity}~\hs{v}, and can now be passed to a \hs{task}. The
result comes as \hs{Identity}~\hs{v}, which we unwrap by \hs{runIdentity}.

@@

\noindent
Here \hs{Const} is a standard Haskell type defined in Fig.~\ref{fig-stdlib}. We
instantiate \hs{f} to \hs{Const}~\hs{[@@k]}. So a value of type \hs{f}~\hs{v},
or in this case \hs{Const}~\hs{[@@k]}~\hs{v}, contains no value \hs{v}, but does
contain a list of keys of type \hs{[@@k]} which we use to record dependencies.
The \hs{fetch} callback that we pass to \hs{task} records a single dependency;
and the standard definition of \hs{Applicative} for \hs{Const} (which we give
in Fig.~\ref{fig-stdlib}) combines the dependencies from different parts of the
task. Running the task with \hs{f}~=~\hs{Const}~\hs{[@@k]} will thus
accumulate a list of the task's dependencies -- and that is just what
\hs{dependencies} does:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies $ fromJust $ sprsh1 "A1"
[@@]
\end{minted}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies $ fromJust $ sprsh1 "B1"
["A1", "A2"]
\end{minted}
\vspace{1mm}

\noindent
Notice that these calls to \hs{dependencies} do no actual computation (in this
case, spreadsheet arithmetic). They cannot: we are not supplying a store or any
input numbers. So, through the wonders of polymorphism, we are able to extract
the dependencies of the spreadsheet formula, and to do so efficiently, simply by
running its code in a different \hs{Applicative}! This is not new, for example
see~\cite{free-applicatives}, but it is extremely cool. We will see a practical
use for \hs{dependencies} when implementing applicative build systems,
see~\S\ref{sec-implementation-make}.

@@

\noindent
This implementation uses the standard Haskell \hs{WriterT} \emph{monad
transformer}~\cite{liang1995monad}, which allows us
to record additional information -- a list of keys of type \hs{[@@k]} -- when
computing a task in an arbitrary monad~\hs{m}. We substitute the given \hs{fetch}
with a \hs{trackingFetch} that, in addition to fetching a value, tracks the
corresponding key. The \hs{task} returns the value of type
\hs{WriterT}~\hs{[@@k]}~\hs{m}~\hs{v}, which we unwrap by \hs{runWriterT}. We
will use \hs{track} when implementing monadic build systems,
see~\S\ref{sec-implementation-shake}; here we show an example of \hs{track}ing
monadic tasks when \hs{m}~\hs{=}~\hs{IO}.

\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ fetchIO k = do putStr (k ++ ": "); read <$> getLine
@\ghci@ track (fromJust $ sprsh2 "B1") fetchIO
C1: 1
B2: 10
(10,["C1","B2"])
\end{minted}
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ track (fromJust $ sprsh2 "B1") fetchIO
C1: 2
A2: 20
(20,["C1","A2"])
\end{minted}
\vspace{1mm}

@@
